---
title: "BIOL540/640: Wildlife Population Genomics"
author: "Elle M. Barnes"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

# Background

In this tutorial, we will simulate a genomic study of Mountain Goats across three isolated peaks: North Range, Central Range, and South Range. We are using 500 single nucleotide polymorphisms (SNPs) to evaluate:

-   Genetic Health (Inbreeding/Heterozygosity)
-   Connectivity (PCA & Isolation by Distance)
-   Adaptation (Fst Outliers/Selection)
-   Individual History (Admixture/Hybridization)

### Learning Outcomes

By the end of this tutorial, students should be able to:

1.  Handled a "raw" data frame
2.  Convert data into a specialized genomic format (genind)
3.  Visualize population-level and individual-level patterns
4.  Identify a specific "conservation-relevant" individual using bioinformatic evidence

### A note about RMarkdown

RMarkdown more-or-less uses the same syntax as general Markdown. Cells that start with '{r}' represent code cells. They are standalone chunks of code that can run with a single click (the green 'play' button in the top right of each code cell). In the space between code cells you can write regular notes just as you would in any other Markdown file.

You can add a new code cell by clicking the **Insert** button in the editor toolbar (top right, a box with a 'C' and a '+' sign) or using the keyboard shortcut: Ctrl + Alt + I (or Cmd + Option + I on macOS).

Outputs from code cells (plots, tables, etc.) will appear inline below the chunk or in the RStudio Viewer pane, depending on your settings.

*To produce your final, formatted Rmd report, click the 'Knit' button in the editor toolbar.* RStudio will process the document, run all the code, and generate an output file (e.g., an HTML file) in the same directory as your .Rmd file. The knit process calls the rmarkdown::render() function internally.

## Load required packages

Run the following code chunk to load the packages we need for the analyses in this tutorial. Note that there are both comment lines (start with '\#') and code. R knows to skip over the comment lines when it executes code, but including them alongside the lines of code themselves can be useful when you first start writing code (your future self will thank past you!).

If you have never used R before, you likely will need to install the packages first before you can load them using the 'library()' function. To do this, un-comment the line containing the 'install.packages()' function.

```{r setup, include=FALSE}
# Initialize the environment
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
# Required libraries
# Run the following line in the console:
# install.packages(c("tidyverse", "adegenet", "hierfstat", "reshape2", "ade4"))

# Load libraries for this tutorial
library(tidyverse)
library(ade4)
library(adegenet)
library(hierfstat)
library(reshape2)
```

# Synthetic Data Generation

We will create genotypes and mock GPS coordinates for our samples. The code below generates this dataset for us simulating a situation where we have sampled 60 individuals and sequenced 500 SNPs.

After running the code chunk below, you will see that it produces a table output of our data (wildlife_df) in the format of a **dataframe**.

### Reminder

While this data is "clean," (it's synthetic after all!), real-world RADseq or Whole Genome Sequencing (WGS) data requires significant **filtering** (e.g., removing SNPs with \>20% missing data) before further analysis.

```{r}
set.seed(42)
n_ind <- 60
n_snps <- 500

# Metadata: Regions and GPS
pops <- factor(c(rep("North", 20), rep("Central", 20), rep("South", 20)))
lat <- c(runif(20, 45, 48), runif(20, 40, 43), runif(20, 35, 38))
long <- c(runif(20, -120, -118), runif(20, -119, -117), runif(20, -118, -116))

# Genotypes: Creating a gradient of allele frequencies to simulate drift/structure
genotypes <- matrix(nrow = n_ind, ncol = n_snps)
for(i in 1:n_snps){
  p_grad <- seq(0.2, 0.8, length.out = 3) 
  genotypes[1:20, i] <- rbinom(20, 2, p_grad[1])
  genotypes[21:40, i] <- rbinom(20, 2, p_grad[2])
  genotypes[41:60, i] <- rbinom(20, 2, p_grad[3])
}

# Combine into dataframe
wildlife_df <- data.frame(ID=1:n_ind, Region=pops, Lat=lat, Long=long, genotypes)
colnames(wildlife_df)[5:504] <- paste0("SNP_", 1:n_snps)

for(i in 1:n_snps){
  p1_allele <- rbinom(1, 1, 0.2) 
  p2_allele <- rbinom(1, 1, 0.8)
  wildlife_df[60, paste0("SNP_", i)] <- p1_allele + p2_allele
}

# Convert to genind object
gen_obj <- df2genind(wildlife_df[, 5:504], pop = wildlife_df$Region, ploidy = 2, ncode = 1)

# Preview the data
head(wildlife_df[,1:10])
```

# 1. Genetic Diversity (Ho vs. He)

Are our populations inbreeding? We can compare Observed ($H_o$) to Expected ($H_e$) heterozygosity to make this determination.

## Mating patterns & genetic processes

*Consider the following:*

-   **Random mating** (also known as 'panmixia'): Any individual in a population has an equal chance of mating with any other individual of the opposite sex, regardless of genotype or phenotype
-   **Inbreeding**: Mating of closely related individuals which increases genetic homozygosity (similarity). It can bring together rare, recessive alleles, often causing harmful health effects known as *inbreeding depression*.
-   **High gene flow**: Transfer of genetic material (alleles) from one population to another through the movement of individuals. It increases genetic diversity within a population but reduces genetic differences between populations, acting as a homogenizing force.

### Understanding statistical results

Error bars represent our confidence in the mean estimate. With 500 SNPs, the error bars will likely be quite small. If we only had 10 SNPs, the bars would be huge, making it hard to tell if populations are actually different.

*Why a paired t-test?*

We use a "paired" test because each SNP has its own unique $H_o$ and $H_e$. We aren't just comparing two random groups of numbers; we are looking at the difference at every single genomic location.

*What does a "significant" result mean?*

If P \< 0.05: We reject the null hypothesis. The difference between $H_o$ and $H_e$ is unlikely to be due to chance. In wildlife, this usually (but not always) points to Inbreeding or Population Bottlenecks. If $H_o$ < $H_e$, this points to Heterozygote Excess which suggests Outbreeding or a Recent Bottleneck (e.g., individuals with different genotypes mate more frequently, higher survival or fitness of heterozygotes (heterozygote advantage), or small population size due to a recent bottleneck which can temporarily increase heterozygotes).

If P \> 0.05: The populations are likely in Hardy-Weinberg Equilibrium. Any small differences we see in the bars are just "sampling noise."

### Question

Q1) Look at the error bars on your Heterozygosity plot. Do the error bars for $H_o$ and $H_e$ overlap? What specific genetic process is likely occurring?

```{r}
# Calculate Per-Locus Stats
stats <- basic.stats(gen_obj) 
  # To interpret this line of code: we are running the function basic.stats() on our dataframe 'gen_obj' and naming that output 'stats'

# Extract Ho and He (Hs) for all loci
# These matrices have loci as rows and populations as columns
ho_raw <- stats$Ho
he_raw <- stats$Hs

# Create a summary table with Means and Standard Errors
# We calculate SE as: sd(x) / sqrt(n)
calc_summary <- function(x) {
  res <- colMeans(x, na.rm = TRUE)
  err <- apply(x, 2, function(col) sd(col, na.rm = TRUE) / sqrt(sum(!is.na(col)))) 
  # hehe I snuck in an apply() function. You are officially a bioinformatician.
  return(list(mean = res, se = err))
} # look you made a custom function!

# Apply your function to our data
sum_ho <- calc_summary(ho_raw)
sum_he <- calc_summary(he_raw)

# 4. Build a dataframe for ggplot
div_summary <- data.frame(
  Region = rep(names(sum_ho$mean), 2),
  Metric = rep(c("Ho", "He"), each = 3),
  Value  = c(sum_ho$mean, sum_he$mean),
  Upper  = c(sum_ho$mean + sum_ho$se, sum_he$mean + sum_he$se),
  Lower  = c(sum_ho$mean - sum_ho$se, sum_he$mean - sum_he$se)
)

# Plot mean and standard error
ggplot(div_summary, aes(x = factor(Region, levels = c("North", "Central", "South")), # this reorders x-axis from alphabetical (default) to spatial
                        y = Value, 
                        fill = Metric)) +
  geom_bar(stat = "identity",position = position_dodge(0.9), alpha = 0.8) +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), 
                width = 0.2, 
                position = position_dodge(0.9)) +
  theme_bw() +
  scale_fill_manual(values = c("#466e95", "#b98f9d")) +
  labs(x = "Region", y = "Heterozygosity Index")+
  theme(legend.text = element_text(size = 10), # the commands in theme() just serve to 'beautify' our plot
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.title.x = element_text(face="bold", size = 14),
        axis.title.y = element_text(face="bold", size = 14),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12)
        )

# But are Ho and He statistically different? 
# Let's perform a Paired t-test for each population...
regions <- levels(wildlife_df$Region)
t_test_results <- list()

for(reg in regions){
  # Extract per-locus values for this specific region
  reg_ho <- ho_raw[, reg]
  reg_he <- he_raw[, reg]
  
  # Conduct the test
  test <- t.test(reg_ho, reg_he, paired = TRUE)
  
  t_test_results[[reg]] <- data.frame(
    Region = reg,
    t_Statistic = round(test$statistic, 3),
    p_Value = round(test$p.value, 5),
    Significant = ifelse(test$p.value < 0.05, "YES", "NO")
  )
}

# Combine and display test results
final_t_table <- do.call(rbind, t_test_results)
print(final_t_table)

```

# 2. Population Structure (PCA)

Let's visualize the broad genetic differences between mountain peaks. To do this, we will perform an ordination known as **Principal Components Analysis (PCA)**. The goal of PCA is to reduce the high-dimensionality (complexity) of our dataset and convert them into a smaller, manageable set of uncorrelated variables called "principal components" while preserving maximum variance.

Below, we will plot only the first two principal components (PC1 and PC2), but know that additional PCs are generated using the prcomp() function. You want to look for clustering of points; the closer two points are to one another in ordination space, the more similar they are genetically (based on the 500 SNPs we sequenced).

### Question

Q2) Does the Central population cluster more closely with the North or the South?

```{r}
# Generate principal components
pca_res <- prcomp(wildlife_df[, 5:504], scale. = TRUE)

# Create a dataframe of just PC1 and PC2
pca_df <- data.frame(pca_res$x[, 1:2], Region = wildlife_df$Region)

# Plot PCA
ggplot(pca_df, aes(x = PC1, 
                   y = PC2, 
                   color = factor(Region, levels = c("North", "Central", "South"))
                   )) +
  geom_point(size = 4, alpha = 0.8) +
  stat_ellipse() +
  theme_bw() +
  labs(color = "Region")+
  scale_color_manual(values = c("#466e95","#E7B800", "#b98f9d")) +
  theme(legend.text = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12)
        )
```

# 3. Isolation by Distance (IBD)

*Does the landscape limit gene flow?* We can correlate genetic distance with physical distance using a Mantel test (a general statistical test for correlations between any two data matrices). We can use the results of our Mantel test to evaluate **isolation by distance** (IBD) where higher geographic distance correlates with higher genetic differentiation. In the case of our mountain goats, the mountains likely prevent goats from dispersing, acting as barriers to gene flow.

### Questions

Q3) What was the p-value of your Mantel Test? Is there a significant relationship between geographic distance and genetic distance?

Q4) If a developer wants to build a resort between the North and South ranges, how might your PCA and Mantel results be used to argue against it?

```{r}
# Calculate distance matrices (using Euclidean distance)
gen_dist_mat <- dist(wildlife_df[, 5:504]) # genetic "distance" 
geo_dist_mat <- dist(wildlife_df[, c("Lat", "Long")]) # geographic distance

# Run the Mantel test
mantel_res <- mantel.rtest(gen_dist_mat, geo_dist_mat, nrepet = 999) # run this analysis across 999 permutations

print(mantel_res)

# Reformat distance data
ibd_df <- data.frame(
  Geo_Dist = as.vector(geo_dist_mat),
  Gen_Dist = as.vector(gen_dist_mat)
)

# Plot correlation
ggplot(ibd_df, aes(x = Geo_Dist, y = Gen_Dist)) +
  geom_point(alpha = 0.3, color = "#466e95") + # alpha modifies the transparency of the points to help us see density
  geom_smooth(method = "lm", color = "red", se = TRUE) + # Adds regression line
  theme_bw() +
  labs(x = "Geographic Distance", y = "Genetic Distance") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(face="bold", size = 14),
        axis.title.y = element_text(face="bold", size = 14)
        )
```

# 4. Selection (Manhattan Plot)

Now, let's scan for "outlier SNPs" that show much higher differentiation than the rest of the genome. We measure this differentiation using the **Fixation Index (**$F_{ST}$). While most of the genome evolves neutrally, some SNPs might be under selection (e.g., genes for cold tolerance in the North). We can identify these by looking for SNPs with unusually high $F_{ST}$ values.

*What is a Manhattan plot?* A Manhattan plot is a specialized scatter plot used in genomics to quickly visualize loci of interest in genome-wide datasets. It displays the statistical significance of thousands of SNPs across chromosomes, where high, "skyscraper-like" peaks indicate strong associations with a specific trait. Since our data is synthetic, we will just plot SNP position in order from 1-500.

### A note about negative Fst

The math behind calculating $F_{ST}$ can produce small negative values when there is more variation *within* a population than between them; for plotting purposes, we usually treat these as 0.

### Questions

Q5) Look at the Manhattan Plot: What might an 'orange' SNP represent in terms of biology?

Q6) Approximately how many SNPs exceeded the 95th percentile threshold (the dashed line)?

Q7) Most SNPs are at the bottom of the graph. Why are only a few SNPs very high?

```{r}
# Calculate per-locus Fst
locus_fst <- stats$perloc$Fst
locus_fst[locus_fst < 0] <- 0 # Correct for sampling noise (i.e., negative Fst values)

# Prepare data for plotting
manhattan_df <- data.frame(Pos = 1:n_snps, Fst = locus_fst)
threshold <- quantile(manhattan_df$Fst, 0.95, na.rm=TRUE) # We'll define outliers as SNPs in the top 5% of Fst values

# Manhattan Plot
ggplot(manhattan_df, aes(x = Pos, y = Fst)) +
  geom_point(aes(color = Fst > threshold)) +
  geom_hline(yintercept = threshold, linetype = "dashed") +
  scale_color_manual(values = c("grey", "orange")) +
  theme_bw() +
  labs(x = "SNP Position", y = "Fixation Index (Fst)")+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(face="bold", size = 14),
        axis.title.y = element_text(face="bold", size = 14)
        )
```

# 5. Ancestry (Structure Plot)

Finally, let's look at the ancestry of each individual goat to find hybrids or migrants. This is another way to assess population structure (as we did with PCA).

In this step, we will use our PCs to identify clusters/ancestry. This is a quick alternative to the software STRUCTURE (a common command-line based bioinformatic pipeline for population genomics).

### Question

Q8) Look at the Admixture Plot: Do you see any goats in the Central range that have North ancestry? What does that suggest about movement?

```{r}
# Get the PCA scores for the individuals
pca_scores <- pca_res$x[, 1]

# Normalize these scores to a 0-1 scale to represent 'Ancestry'
# We treat the extremes (North and South) as the 'ancestral' poles
min_score <- min(pca_scores)
max_score <- max(pca_scores)
ancestry_estimate <- (pca_scores - min_score) / (max_score - min_score)

# Create a dataframe where each individual has two 'ancestral' components
plot_df <- data.frame(
  ID = 1:n_ind,
  Region = wildlife_df$Region,
  Ancestry_A = ancestry_estimate,
  Ancestry_B = 1 - ancestry_estimate
)

# Reformat the data
plot_df_long <- plot_df %>%
  pivot_longer(cols = starts_with("Ancestry"), names_to = "Cluster", values_to = "Prob")

plot_df_long <- within(plot_df_long, Region <- factor(Region, levels = c('North', 'Central', 'South')))

# Structure plot
ggplot(plot_df_long, aes(x = factor(ID), y = Prob, fill = Cluster)) +
  geom_bar(stat = "identity", width = 1) +
  facet_grid(~Region, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = c("#466e95", "#b98f9d")) +
  theme_bw() +
  labs(x = "Individual Goat", y = "Ancestry %", fill = "Cluster") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(face="bold", size = 12),
        axis.title.y = element_text(face="bold", size = 12)
        )
```

# 6. The Mystery Individual

You might have noticed, but when we first create the dataset, I added in a mystery individual with a 'different' genetic history. Can you determine which individual this is?

### Questions

Q9) Who is the mystery individual?

Q10) Based on their ancestry, what is their likely 'origin story'?

Q11) What is the conservation implication of finding this mystery goat? Does the existence of this individual change your view on how "isolated" the South population really is? Why? 

# Final Thoughts

Q12) Imagine you are the Lead Biologist for the Mountain Goat Recovery Team. Based on today's lab, rank the three populations from Highest Conservation Priority to Lowest, and give one reason why.

